---
author: "Jette Steinbach"
date: "16/02/2022"
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```
# LTFHpp

LTFHpp extends the package [LTFHPlus](https://emilmip.github.io/LTFHPlus/) by Emil M. Pedersen
to allow for more flexible family structures. 
As LTFHPlus, LTFHpp can be used to estimate an individual's genetic component of the full liability, the full liability or both by accounting for the family history and population prevalences. 
LTFHpp inherits its efficiency from LTFHPlus, as it uses the same methods, i.e. Gibbs sampler implemented with [Rcpp](http://www.rcpp.org/). 
A detailed description of the liability threshold model conditioned on family history, age of onset and sex can be found [here](https://doi.org/10.1016/j.ajhg.2022.01.009).


# Installation

You can install LTFHpp by:

```{r eval=FALSE}
devtools::install_github("JetteS/LTFHpp")
```

<!-- badges: start -->
[![R build status](https://github.com/JetteS/LTFHpp/workflows/R-CMD-check/badge.svg)](https://github.com/JetteS/LTFHpp/actions)
<!-- badges: end -->

