---
title: "From trio information to full families"
author: "Emil M. Pedersen"
date: "`r Sys.Date()`"
output: html_document
---

```{r message=FALSE, warning=FALSE}
library(LTFHPlus)
library(dplyr)
library(igraph)
library(kinship2)
```


   In this document, we will present how we can go from trio information to full families that can be used to calculate kinship matrices. By trio information, we specifically mean knowing the id of the child and the id of the child's mother and father. Kinship matrices are essential when estimating the liabilities with the `estimate_liability()` function of the package. This addition help with the process of identifying related individuals and subsequent construction of the kinship matrix.

### From trio information to graph
   The trio information can be used to create extended families manually by first identifying parents, grandparents, great-grandparents, etc.. From there, siblings, aunts and uncles, cousins, etc.. can also be identified. However, this is a tedious process and it is easy to miss family members.
   We have developed a function that can find all family member that are related of degree $n$ or closer that does not rely on the tedious process of identifying each family role manually.


Below is an example data set of a family. It contains half-siblings, half-aunts and -uncles, as well as cousins and individuals that have married into the family. An example is *mgm* meaning *maternal grandmother*, *hspaunt* meaning *paternal half-aunt*, or *hsmuncleW* meaning *maternal half-uncle's wife*.

```{r}
family = tribble(
  ~id, ~momcol, ~dadcol,
  "pid", "mom", "dad",
  "sib", "mom", "dad",
  "mhs", "mom", "dad2",
  "phs", "mom2", "dad",
  "mom", "mgm", "mgf",
  "dad", "pgm", "pgf",
  "dad2", "pgm2", "pgf2",
  "paunt", "pgm", "pgf",
  "pacousin", "paunt", "pauntH",
  "hspaunt", "pgm", "newpgf",
  "hspacousin", "hspaunt", "hspauntH",
  "puncle", "pgm", "pgf",
  "pucousin", "puncleW", "puncle",
  "maunt", "mgm", "mgf",
  "macousin", "maunt", "mauntH",
  "hsmuncle", "newmgm", "mgf",
  "hsmucousin", "hsmuncleW", "hsmuncle"
)

thrs =  tibble(
 id = family %>% select(1:3) %>% unlist() %>% unique(),
 lower = sample(c(-Inf, 2), size = length(id), replace = TRUE),
 upper = sample(c(2, Inf), size = length(id), replace = TRUE),
 sex = case_when(
   id %in% family$momcol ~ "F",
   id %in% family$dadcol ~ "M",
   TRUE ~ NA)) %>% 
  mutate(sex = sapply(sex, function(x) ifelse(is.na(x), sample(c("M", "F"), 1), x)))
```

The object `family` is meant to represent the trio information that can be found in registers. It is possible to have multiple families in the same input data or single individuals with no family links.


```{r}
graph = prepare_graph(.tbl = family, 
                      thresholds = thrs,
                      fcol = "dadcol",
                      mcol = "momcol",
                      icol = "id")
graph
```

The object `graph` is a directed graph constructed from the trio information in `family` and is build using the ***igraph*** package. The direction in the graph is from parent to offspring. 

### From graph to subgraph and kinship matrix

We can construct a kinship matrix from all family members present in `family`, or we can consider only the family members that are of degree $n$. We can identify the family members of degree $2$ like this:

```{r, fig.alt = "Plot of the identified pedigree. Pedigree plotted with igraph package."}
# make_ego_graph returns list, even for node input of length 1
fam_graph = make_ego_graph(graph = graph, 
                           order = 2,
                           nodes = "pid")[[1]]
plot(fam_graph, layout = layout_as_tree,
     vertex.size = 27.5,
     vertex.shape = "rectangle",
     vertex.label.cex = .75,
     edge.arrow.size = .3) 
```

  In particular, individuals such as paternal uncle's child (i.e a cousin, coded as pucousin above) is not present with this relatedness cut-off as such family members are of degree $3$. 

### Calculate kinship matrix

Finally, the kinship matrix can be calculated with `get_kinship()` (output made nicer with round) in the following way:
```{r}
# the kinship matrix is multiplied by 100 and rounded for illustrative purposes!
round(get_kinship(fam_graph, h2 = 1, index_id = "pid", add_ind = FALSE) * 100, 2)
```


A function called `graph_to_trio()` has been included in the package, which can convert from the graph object back into a trio object. This function is useful if you want to use the functionality of other packages that rely on trio information. One such example is using the plotting functionality of pedigrees in kinship2.

```{r}
trio = graph_to_trio(graph = fam_graph, fixParents = TRUE)
trio
```
which can be used to utilise the powerful plotting tool kit available in the kinship2 package.
```{r, fig.alt = "Plot of the identified pedigree. Pedigree plotted with kinship2 package."}
pedigree = with(trio,kinship2::pedigree(id = id, dadid = dadid,momid =  momid,sex =  sex))

plot(pedigree)
```