---
title: "LT-FH++ Example"
author: "Emil M. Pedersen"
date: "5/7/2021"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{LT-FH++ Example}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
	echo = TRUE,
	message = TRUE,
	warning = TRUE
)
```

First, we will load the packages that we will need for the example:
```{r message=FALSE}
library(LTFHPlus)
library(dplyr)
library(MASS)
library(future.apply)
library(tidyr)
```

### Global parameters
The strategy from here is to generate the necessary information for each individual in the example. We will start out by setting some global parameters:

```{r}
N = 100 #number of individuals
h2 = .5 #heritability
nsib = 2 #number of siblings

#calculates the sex-specific prevalences used to determine status:
multiplier = 1
prev = c(0.08, 0.02) * multiplier

nthreads = 5  # number of threads to use for LT-FH++
```
It is worth noting that these values can be set more or less anything the reader would want/need. Our choice of prevalence vector is to illustrate LT-FH++'s ability to estimate the genetic liability, even when the cases are not equally distributed between the sexes. Other prevalence vectors could be used, for instance one with the same prevalence between sexes.

### Setting up input

LT-FH++ requires 3 types of input, name **h2**, **phen**, and **thr**, which is the heritability on liability scale, a list of ids (where each list entry is a family), and finally the upper and lower limit for each individual used by the truncated normal in LT-FH++.

With the global parameters set, we will simulate the liabilities for each family member based on the multivariate normal described in the [LT-FH++] paper. Next, we assign ages to each individual. The child's age act as a baseline age, and the age of parents and siblings are determined based off of the child's age.
```{r}
#simulate liabilities
liabs = MASS::mvrnorm(n = N, mu = rep(0, 4 + nsib),
                      Sigma = construct_covmat(fam_vec = c("m", "f", if (nsib > 0) paste0("s", 1:nsib)),
                                               sq.herit = h2)
                      )

#Age of children, parents, and siblings
child_age = runif(N, 15, 60)
father_age  = child_age + runif(N, 20, 35)
mother_age  = child_age + runif(N, 20, 35)
children_sex = sample(1:2, size = N, replace = TRUE)
if(nsib > 0) {
  sibling_sex = replicate(nsib, sample(1:2, size = N, replace = TRUE))
  sibling_age = replicate(nsib, runif(N, -15, 15) + child_age)
}
```
#### Generating phen
Next, we want to assign sex, status, age-of-onset(aoo) to each individual, including the family members.  

```{r message=FALSE}
# Constructing tibble with list entries,
# each row is a family. assuming order: os, father, mother, siblings (if any)
phen = lapply(1:N, function(i) {
  #get full liabilities of all family members
  cur_liabs  = liabs[i, 1:(4 + nsib)][-1] 
  # assign ids
  id_vec     = c(i , paste0(i, c("_f", "_m")), if(nsib > 0) paste0(i, "_s", 1:nsib) ) 
  #assign sex
  sex_vec    = c(children_sex[i], 1:2, if(nsib > 0) sibling_sex[i,]) 
  # assign age
  age_vec    = c(child_age[i], father_age[i], mother_age[i], if (nsib > 0) sibling_age[i,]) 
  # assign status
  status_vec = sapply(seq_along(sex_vec), function(ii) { 
    (cur_liabs[ii] > qnorm(prev[sex_vec[ii]], lower.tail = F)) + 0L
  })

  #assign age-of-onset for cases
  aoo_vec = rep(NA, length(id_vec)) 
  
  cases = status_vec == 1
  if(any(cases)) {
    aoo_vec[cases] = pmax(convert_liability_to_aoo(liability = cur_liabs[cases],
                                                   pop_prev = prev[sex_vec[cases]]),
                          4)
  }
  #return information
  c(list(id_vec),
    list(sex_vec),
    list(status_vec),
    list(age_vec),
    list(aoo_vec))
}) %>% do.call("rbind", . ) %>% #outputs list, combining with rbind 
  as_tibble(.name_repair = 'unique') %>% # converting to tibble
  rename("ids" = "...1",
         "sex" = "...2", 
         "status" = "...3", 
         "age" = "...4", 
         "aoo" = "...5") %>% # renaming columns
  print()
```
The output is a tibble with list entries, meaning each row is a family in the order of child, father, mother, and siblings (if any). This keeps all relevant information for each family accessible in an easy manner for processing later on. Each family can be easily accessed, e.g.:

```{r }
phen %>% 
  slice(1) %>% 
  tidyr::unnest(cols = colnames(phen))
```
Since we expect the family sizes to vary between families and we wanted to accomodate this, we found it simplest to work with lists or tibbles with list entries. This means the **phen** input, which is the first input, is simply the *ids* column from *phen*.

#### Generating thr

The other input is **thr**, which will contain the ids and an upper and lower limit of possible liabilities for each individual. 
```{r }
#assigning thresholds for each family
thr = lapply(1:nrow(phen), function(n) {
  res = list()
  ids   = phen$ids[[n]]
  cases = phen$status[[n]] == 1
  lower = rep(-Inf, length(ids))
  upper = rep( Inf, length(ids))
  
  if(any(cases)) {
    lower[cases] <- upper[cases] <- convert_age_to_thresh(age = phen$aoo[[n]][cases], 
                                                          pop_prev = prev[phen$sex[[n]][cases]])
  }
  
  upper[!cases] <- convert_age_to_thresh(age = phen$age[[n]][!cases],
                                         pop_prev = prev[phen$sex[[n]][!cases]])
  res$ids = ids
  res$lower = lower
  res$upper = upper
  res
}) %>% 
  lapply(., as_tibble) %>% 
  bind_rows() %>%
  print()

```

### Running LT-FH++

With all input parameters generated and the heritability set in the global settings, we have everything we need to run *LT-FH++*. The package **Future.apply** can be used to parallelize the computations for LT-FH++, and we highly recommend doing this, since all modern computers have more than 2 cores available to them. The future framework also allows for easy use on high performance computing (HPC) clusters. In this example, we will utilize `r nthreads` threads. 

```{r }
phen = dplyr::select(phen, ids)

#Setting up parallelization backend
plan(tweak(multisession, workers = nthreads))
#performs LT-FH++ analysis
simu_liab = estimate_gen_liability(h2 = h2,
                       phen = phen,
                       thr = thr,
                       id_col = "ids")
simu_liab = simu_liab %>% mutate(ids = sapply(ids, function(x) x[1])) %>% print()
```
From here the object simu_liab can be saved and used in your GWAS method of choice.

