---
title: "LT-FH++ Example No Family History"
author: "Emil M. Pedersen"
date: "7/7/2021"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{LT-FH++ Example No Family History}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(
	echo = TRUE,
	message = TRUE,
	warning = TRUE
)
```

First, we will load the packages that we will need for the example:
```{r message=FALSE}
library(LTFHPlus)
library(dplyr)
library(future.apply)
```

### Global parameters
The strategy from here is to generate the necessary information for each individual in the example. We will start out by setting some global parameters:

```{r}
N = 10000 #number of individuals
h2 = .5 #heritability

#calculates the sex-specific prevalences used to determine status:
multiplier = 1
prev = c(0.08, 0.02) * multiplier

nthreads = 5  # number of threads to use for LT-FH++ with no FH
```
It is worth noting that these values can be set more or less anything the reader would want/need. Our choice of prevalence vector is to illustrate LT-FH++'s ability to estimate the genetic liability, even when the cases are not equally distributed between the sexes. Other prevalence vectors could be used, for instance one with the same prevalence between sexes.

### Setting up input

Simulating liabilities with no family history is slightly easier, since the multivariate normal only has two entries, one for the genetic and one for the full liability.

```{r}
#covariance matrix with no FH
cov_mat = matrix(h2, ncol = 2, nrow = 2)
cov_mat[2,2] = 1


#simulate liabilities & make tibble
liabs = MASS::mvrnorm(n = N, mu = rep(0, 2), Sigma = cov_mat) %>%
  as_tibble(.name_repair = 'unique') %>%
  rename("child_gen"  = "...1",
         "child_full" = "...2")
```

#### Generating phen

When we want to estimate the genetic liability for en individual without any family history, we only need the ***phen*** object. In short, ***phen*** will contain the ids and upper and lower limit for the full liability. 

```{r}
#construct phen
phen = tibble(
  ids = 1:N,
  child_gen  = liabs$child_gen,
  child_full = liabs$child_full,
  age  = runif(N, 10, 60),
  sex        = sample(1:2, size = N, replace = T), # 1 and 2 to avoid "+1" through out the example
  status     = (child_full > qnorm(prev[sex], lower.tail = F)) + 0L,
  lower      = -Inf,
  upper      = Inf
) 

#who are the cases?
cases = which(phen$status == 1) 
#fill in aoo
phen$aoo = NA
phen$aoo[cases] = LTFHPlus::liab_to_aoo(liab = phen$child_full[cases], pop_prev = prev[phen$sex[cases]])
#fill in upper and lower limit
phen$upper[-cases] = LTFHPlus::age_to_thres(age      = phen$age[-cases], 
                                            pop_prev = prev[phen$sex[-cases]])
phen$lower[cases]  = LTFHPlus::age_to_thres(age      = phen$aoo[cases], 
                                            pop_prev = prev[phen$sex[cases]])

# keep only relevant information
phen = dplyr::select(phen, ids, status, lower, upper, child_gen) %>% 
  print()

```
***phen*** contains the prevalence information, e.g. age & sex, needed to find the upper and lower limit. Strictly speaking, only the ids, lower and upper limit is necessary from this point on.


### Running LT-FH++ no family history

With all input parameters generated and the heritability set in the global settings, we have everything we need to run *LT-FH++*. The package **Future.apply** can be used to parallelize the computations for LT-FH++, and we highly recommend doing this, since all modern computers have more than 2 cores available to them. The future framework also allows for easy use on high performance computing (HPC) clusters. In this example, we will utilize `r nthreads` threads. 

```{r}
#start parallelization
plan(tweak(multisession, workers = nthreads))

ltfhpp_nofh = estimate_gen_liability_noFH(phen = phen, h2 = h2)
ltfhpp_nofh
```