---
title: "LT-FH++ Example"
author: "Emil M. Pedersen"
date: "6/9/2022"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{LT-FH++ Example}
  %\VignetteEngine{knitr::knitr}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
	echo = TRUE,
	message = TRUE,
	warning = TRUE
)
	h2 = .5
	nthreads = 4
```

First, we will load the packages that we will need for the example:
```{r message=FALSE}
library(LTFHPlus)
library(dplyr)
library(MASS)
library(future.apply)
library(tidyr)
```

For simplicty's sake, we will use the same prevalence for both sexes. However, it is worth noting that in real-world applications, the cumulative incidence proportions needed for the thresholds, will rarely be identical for both sexes.   

### Setting up input

Single trait LT-FH++ requires 3 types of input, namely the liability-scale heritability, the family relationships and the status, age or age of onset, sex, and birth year for each family member, and finally cumulative incidence proportions (CIPs). The more detailed cumulative incidence proportions, the better. In the LT-FH++ paper, we used CIPs that were stratified by birth year and sex and fixed the upper and lower liability threshold for cases to $\Phi(1 - CIP)$, since it provided a very accurate estimate for the full liability. Each CIP curve was a function of age, where age represented the age of onset or the current age of a control. If less detailed information, such as CIPs stratified by sex, but not birth year, we recommend setting the lower threshold to be $\Phi(1 - CIP)$ and the upper limit to infinity for cases. With this information, the thresholds needed for the age-dependent liability threshold model can be assigned. We have provided a function called `prepare_LTFHPlus_input` that can help users convert the input into a suitable format. An example of how the input may look and how to use the function can be found in *From CIP and family to LT-FH++ input*. 


### Generating phenotypes

We have implemented a function that allows users to simulate under the liability threshold model for a given family structure, i.e. mother, father, sibling, or sibling1, sibling2, child1, child2, etc. The function simply needs a family structure, heritablity, and population prevalence.For simplicity's sake, we will use the same prevalence for both sexes. However, it is worth noting that in real-world applications, the cumulative incidence proportions needed for the thresholds, will rarely be identical for both sexes. The function `simulate_under_LTM` outputs a list with two tibbles. The first one `sim_obs` contains the underlying liabilities, status, and age or age of onset for each family and family members. The second tibble `thresholds` contains the formatted input that is ready to be input to estimate the genetic liability.  

```{r}
sims <- simulate_under_LTM(fam_vec = c("m","f","s1"), 
                           n_fam = NULL, 
                           add_ind = TRUE, 
                           h2 = h2, 
                           n_sim = 100, 
                           pop_prev = .1)
sims
```
`thresholds` keeps track of families and family members with `fam_id` and `indiv_id`. Here we simply use dummy variables for both. In real-world data, the `indiv_id` is usually some pseudonymized identifier unique to each individual. There is total freedom to set `fam_id` to whatever, as long as it is unique to each family. One choice could be the pseudonymized identifier for the index person in a family, or simply 1 to the total number of families. Next, `role` identifies each family member's relationship to the index person. They follow a system shown in the documentation for `construct_covmat`. If a family member does not fit one of the shown, a suitable covariance matrix cannot be constructed. If more detailed family structures are needed, please contact one of the maintainers of <tt>LTFHPlus</tt>. Finally, the lower and upper liability thresholds are provided for each family member. Previously, we utilised list of lists in <tt>R</tt> to link information to the correct individuals. Going forward, we will still support this format, but in order to ease the use of non-<tt>R</tt> users, we have opted for a long format, where each row is an indivdual instead of a family, such that the input can be generated with the user's preferred software. 


### Running LT-FH++

With all input parameters generated and a chosen heritability, we have everything we need to run LT-FH++. We highly recommend the package **Future** to parallelize the computations for LT-FH++. Since all modern computers have more than 2 cores available to them, it should be available to all users. If a user is not able to parallelize, then <tt>LTFHPlus</tt> can still be used, but it will be far slower. The future framework also allows for easy use on high performance computing (HPC) clusters. In this example, we will utilize `r nthreads` threads. 

```{r }
#Setting up parallelization backend
plan(tweak(multisession, workers = nthreads))
#performs LT-FH++ analysis
simu_liab = estimate_liability(.tbl = sims$thresholds,
                               h2 = h2,
                               pid = "indiv_ID",
                               fam_id = "fam_ID",
                               role = "role")
simu_liab
```
From here the object `simu_liab` can be saved and used in your GWAS method of choice.

